Structuring Theories with Implicit Morphisms

### 2019-02-08 appl.tex
minor comments by reviewers

\paragraph{Renamings}
We say that a named morphism $r:S\to T=\{\ldots\}$ is a \textbf{renaming} if
\begin{compactitem}
 \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
 \item every such $T$-constants $c'$ occurs in exactly one assignment.
\end{compactitem}
Clearly, every renaming is an isomorphism.
The inverse morphisms contains the flipped assignments $c':=c$.

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
  \item This is allowed if there are no implicit morphism between $s$ and $t$ yet.
  \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
\end{compactitem}

@@

\paragraph{Definitional Extensions}
We say that the named theory $T$ is a \textbf{definitional extension} of $S$ if $T=S$ or the body of $T$ contains
\begin{compactitem}
 \item only constant declarations with definiens,
 \item only include declarations of theories that are definitional extensions of $S$.
\end{compactitem}

If $T$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves. In particular, the isomorphism $S\to T$ maps $S$-constants to themselves and expands the definiens of all other constants.

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{S}$ of a named theory $S$ inside theory $T$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{S}\in\IMo{T}{S}$ (in addition to the implicit morphism $\id{T}\in\IMo{S}{T}$ that is induced by the inclusion).
\end{compactitem}

@@

\paragraph{Canonical Isomorphisms}
If we have isomorphisms $m:s\to t$ and $n:t\to s$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
This requires no language extensions.

@@

We define the declaration $M=N:S\to T \keyword{by} \{\sigma\}$ to be well-formed iff
\begin{compactitem} 
  \item $M:S\to T$ and $N:S\to T$ are well-formed morphisms
  \item $\sigma$ contains exactly one assignment $c:=p$ for every $(c:E)\in\flt{S}$
  \item for each of these assignments $c:=p$, the term $p$ is a proof of $\vdash_T M(c)=N(c)$.
\end{compactitem}

@@

A major drawback of using modular theories is that it can preclude transparent refactoring.
For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $r$ to $s$.
This change to should be straightforward as it does not change the semantics of $t$.

However, this is not a local change.
It also requires updating every qualified reference from $r?n$ to $s?n$.
Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know or does not have access to.

>>>>

\paragraph{Renamings}
We say that a named morphism $r:S\to T=\{\ldots\}$ is a \textbf{renaming} if
\begin{compactitem}
 \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
 \item every  $T$-constant $c'$ without definiens occurs in exactly one assignment.
\end{compactitem}
Clearly, every renaming is an isomorphism.
The inverse morphisms contains the flipped assignments $c':=c$.

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
  \item This is allowed if there are no implicit morphisms between $s$ and $t$ yet.
  \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
\end{compactitem}

@@

\paragraph{Definitional Extensions}
We say that the named theory $T$ is a \textbf{definitional extension} of $S$ if $T=S$ or the body of $T$ contains only
\begin{compactitem}
 \item constant declarations with definiens, and
 \item include declarations of theories that are definitional extensions of $S$.
\end{compactitem}

If $T$ is a definitional extension of $S$, it is easy to prove that $T$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves. In particular, the isomorphism $T\to S$ maps $S$-constants to themselves and expands the definiens of all other constants.

We make the following extension to syntax and semantics:
\begin{compactitem}
  \item An include declaration $\icl{S}$ of a named theory $S$ inside a theory $T$ may carry the attribute \keyword{definitional}.
  \item In that case, we define $\id{S}\in\IMo{T}{S}$ (in addition to the implicit morphism $\id{T}\in\IMo{S}{T}$ that is induced by the inclusion).
\end{compactitem}

@@

\paragraph{Canonical Isomorphisms}
If we have isomorphisms $m:S\to T$ and $n:T\to S$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
This requires no language extensions.

@@

We define the declaration $\keyword{equal} M=N:S\to T \keyword{by} \{\sigma\}$ to be well-formed iff
\begin{compactitem} 
  \item $M:S\to T$ and $N:S\to T$ are well-formed morphisms
  \item $\sigma$ contains exactly one assignment $c:=p$ for every $(c:E)\in\flt{S}$
  \item for each of these assignments $c:=p$, the term $p$ is a proof of $\vdash_T M(c)=N(c)$.
\end{compactitem}

@@

A major drawback of using modular theories is that it can preclude transparent refactoring.
For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
This change should be straightforward as it does not change the semantics of $t$.

However, this is not a local change.
It also requires updating every qualified reference from $s?n$ to $r?n$.
Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know or does not have access to.

### 2019-02-08 implicit.tex
minor comments by reviewers

Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
It contains all theories but only some of the morphisms --- the ones designated as \emph{implicit}.
Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ --- if this morphism is $i$, we write $\ipc{i}{S}{T}$.
%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
%(This is well-defined because composition is associative and identity is neutral.)
%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.
The implicit-diagram generalizes the inclusion relation:
All identity and inclusion morphisms are implicit, and we recover the inclusion relation $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
And just like inclusion, the relation ``exists $i$ such that $\ipc{i}{S}{T}$'' is a preorder.

@@

Adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
\end{definition}

@@

\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
For include morphisms, we add the following definition: if theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.

Combined with composition morphisms, we see that all transitive includes between theories are implicit.
That corresponds to the intuition that anything that is included is available by its original name.
\end{example}

\begin{example}
	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
	For example, we can write $a/b$ in $\cn{Group}$, where $/$ refers to the identifier $\cn{DivGroup}?\cn{div}$.
\end{example}

@@

The following theorem is our central theoretical result.
It shows that the modification made in Def.~\ref{def:useimpl} has the intended properties:
\begin{theorem}[Conservativity of Implicit Morphisms]
For well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$:
\begin{compactenum}
 \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
 \item Whenever the modified system proves $\vdash_T e=e':E$, then the original system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
\end{compactenum}
Here $\ov{E}$ is the expression that arises from $E$ by recursively replacing every constant with its definiens.
\end{theorem}

@@

For the second claim, we proceed by induction on derivations.
We only need to consider the case where the modified rule was applied.
So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
We distinguish two cases:
\begin{compactitem}
 \item $M(c)=c$, i.e., $e$ is absent, and $\ov{c}=c$.
   According to the definition of $\flt{S}$, this is only possible if $S\harr T$ (including the special case $S=T$).
   In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
   Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
 \item $M(c)\neq c$, i.e., $e$ is present or $M$ assigns a non-trivial value to $c$.
  Definition expansion eliminates $c$ in favor of $M(c)$, and thus $\ov{c}=\ov{M(c)}$.
  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
\end{compactitem}
\end{proof}

>>>>

Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit diagram}.
It contains all theories but only some of the morphisms --- the ones designated as \emph{implicit}.
Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ --- if this morphism is $i$, we write $\ipc{i}{S}{T}$.
%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
%(This is well-defined because composition is associative and identity is neutral.)
%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.
The implicit-diagram generalizes the inclusion relation:
All identity and inclusion morphisms are implicit, and we recover the inclusion relation $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
And just like inclusion, the relation ``exists $i$ such that $\ipc{i}{S}{T}$'' is a preorder.

@@

Adding an $\keyword{implicit}$ attribute to a declaration is well-formed only if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
\end{definition}

@@

\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
For include morphisms, we add the following definition: if a theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.

Combined with composition morphisms, we see that all transitive includes between theories are implicit.
That corresponds to the intuition that anything that is included is available by its original (i.e. qualified) name.
\end{example}

\begin{example}
	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
	For example, we can write $a/b$ in $\cn{Group}$, where $/$ refers to the identifier $\cn{DivGroup}?\cn{div}$.

	(Note that notations are carried over by implicit morphisms as well)
\end{example}

@@

The following theorem is our central theoretical result.
It shows that the modification made in Def.~\ref{def:useimpl} has the intended properties:
\begin{theorem}[Conservativity of Implicit Morphisms]
For well-formed diagrams and $S,T\in\Thy$ and $M\in\IMo{S}{T}$:
\begin{compactenum}
 \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
 \item Whenever the modified system proves $\vdash_T e=e':E$, then the original system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
\end{compactenum}
Here $\ov{E}$ is the expression that arises from $E$ by recursively replacing every constant with its definiens.
\end{theorem}

@@

For the second claim, we proceed by induction on derivations.
We only need to consider the case where the modified rule was applied.
So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
We distinguish two cases:
\begin{compactitem}
 \item $M(c)=c$, i.e., $e$ is absent, and $\ov{c}=c$.
   According to the definition of $\flt{S}$, this is only possible if $S\harr T$ (including the special case $S=T$).
   In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
   Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
 \item $M(c)\neq c$, i.e., $e$ is present or $M$ assigns a non-trivial value to $c$.
  Definition expansion eliminates $c$ in favor of $M(c)$, and thus $\ov{c}=\ov{M(c)}$.
  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$. That follows from the judgment preservation of morphisms.
\end{compactitem}
\end{proof}

### 2019-02-08 intro.tex
minor comments by reviewers

\paragraph{Motivation}
Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps,isabelle_locales}, and knowledge representation \cite{RK:mmt:10}.
Closely related concepts are used in both object-oriented (\emph{classes}) and (\emph{type classes}) functional programming languages.

@@

For example, a document might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.%
\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$. Models are translated in the opposite direction.}
%Crucially, $m(-)$ preserves typing and provability.

The key idea behind implicit morphisms is very simple:
We maintain an additional diagram $I$, which is commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions, with the semantics being that $c$ abbreviates $i(c)$.
For example, in the diagram above, we may choose to label \cn{DG2G} implicit.
Immediately, every abbreviation or theorem that we have formulated in the theory \cn{DivGroup} becomes available for use in \cn{Group} without any syntactic overhead.
We can even label \cn{G2DG} implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that \cn{Group} and \cn{DivGroup} are just different formalizations of the same concept.
While these morphisms must be labeled manually, any inclusion morphism like the one from \cn{Monoid} to \cn{Group} can be made implicit automatically.

\paragraph{Contribution}
At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
We recommend using implicit morphisms in all theory morphism--based formalisms.
More concretely, we present a formal system for developing structured theories with implicit morphisms.
Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
\mmt is logic-independent, i.e., allows embedding a large variety of declarative languages (logics, type-theories, etc.).
Therefore, all our results can be directly applied to any language $L$ represented in \mmt or easily transferred to dedicated implementations of $L$.

>>>>

\paragraph{Motivation}
Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps,isabelle_locales}, and knowledge representation \cite{RK:mmt:10}.
Closely related concepts are used in both object-oriented (\emph{classes}) and functional (\emph{type classes}) programming languages.

@@

For example, a document might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.%
\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to a $T$-expression $m(E)$. Models are translated in the opposite direction.}
%Crucially, $m(-)$ preserves typing and provability.

The key idea behind implicit morphisms is very simple:
We maintain an additional diagram $I$, which is a commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions, with the semantics being that $c$ abbreviates $i(c)$.
For example, in the diagram above, we may choose to label \cn{DG2G} implicit.
Immediately, every abbreviation or theorem that we have formulated in the theory \cn{DivGroup} becomes available for use in \cn{Group} without any syntactic overhead.
We can even label \cn{G2DG} implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that \cn{Group} and \cn{DivGroup} are just different formalizations of the same concept.
While these morphisms must be labeled manually, any inclusion morphism like the one from \cn{Monoid} to \cn{Group} can be made implicit automatically.

\paragraph{Contribution}
At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
We recommend using implicit morphisms in all theory--morphism--based formalisms.
More concretely, we present a formal system for developing structured theories with implicit morphisms.
Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
\mmt is logic-independent, i.e., allows embedding a large variety of declarative languages (logics, type-theories, etc.).
Therefore, all our results can be directly applied to any language $L$ represented in \mmt or easily transferred to dedicated implementations of $L$.

### 2019-02-08 mmt.tex
minor comments by reviewers

\paragraph{Flat Modules}
\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$.
Every theory induces the set $\Exp{T}$ for the set of closed \textbf{expressions} using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

@@

A \textbf{diagram} consists of a set of theory and morphism declarations.
\begin{nomodexp}
For a given diagram, we write $\Thy$ for the set of declared \textbf{theories}.
And we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
\begin{compactitem}
 \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
 \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
 \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
\end{compactitem}
$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
\end{nomodexp}
\begin{modexp}
Every diagram induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
\end{modexp}

@@

\begin{example}\label{ex:thgroup}
For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.

@@

\begin{definition}[Flattening]
For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
 \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
Here $\fltd{T}{D}$ is the flattening of declaration $D$ relative to $\flt{\Sigma}$.
At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

@@

\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
We do this as follows
  \[\fltd{\Sigma}{(\icl{S})} = \flt{S}\]
This has the effect of copying over all declarations from the included into the including theory.

Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
Therefore, includes can never lead to name clashes.
Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.

The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening is not always defined:
  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
and $\sigma,\icl{M}$ is well-formed $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
\end{example}

>>>>

\paragraph{Flat Modules}
\textbf{Flat theories} are lists of constant declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
We write $\dom{T}$ for the set of constant identifiers $c$ in $T$.
Every theory induces the set $\Exp{T}$ for the set of closed \textbf{expressions} using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

@@

A \textbf{diagram} consists of a set of theory and morphism declarations.
\begin{nomodexp}
For a given diagram, we write $\Thy$ for the set of declared \textbf{theories}.
For theories $S,T\in\Thy$, we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
\begin{compactitem}
 \item for every declaration $m:S\to T=\{\cdot\}$, we have $m\in\Mo{S}{T}$,
 \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
 \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$ (where $M;N$ corresponds to the usual function composition).
\end{compactitem}
$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
\end{nomodexp}
\begin{modexp}
Every diagram induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
\end{modexp}

@@

\begin{example}\label{ex:thgroup}
For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
\texttt{type} is the universe of types, \lstinline|A|$\rightarrow$\lstinline|B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
We also use \mmt notations, which are attached to constant declarations as \lstinline|# <notation>|; these are omitted from the formal grammar above because we only use them in the examples.

@@

\begin{definition}[Flattening]
For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
 \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D},\]
 where $\cdot$ denotes the empty sequence.

Here $\fltd{T}{D}$ is the flattening of declaration $D$ relative to $\flt{\Sigma}$.
At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

@@

\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
The $\keyword{include}$ operator adds new declarations $D$ to a theory, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
We do this as follows
  \[\fltd{\Sigma}{(\icl{S})} = \flt{S}\]
This has the effect of copying over all declarations from the included into the including theory.

Note that the definition of $\fltd{\Sigma}{(n:E=e)}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
Therefore, includes can never lead to name clashes.
Also note that, because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.

The situation is slightly more complicated for morphisms: if a morphism $\sigma$ out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening is not always defined:
  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
if $\sigma,\icl{M}$ is well-formed and $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
\end{example}

### 2019-02-08 paper.tex
further cleanup, conclusion still missing
cleaned up intro and conc
finished for submission
minor comments by reviewers

@@

\usepackage{macros/sembracket}
\usepackage{macros/mytikz}
\usepackage{macros/basics}
%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
\usepackage{local}

@@

\begin{abstract}
We introduce \emph{implicit} morphisms as a concept in formal systems based on theories and theory morphisms.
The idea is that there may be at most one implicit morphism from a theory $S$ to a theory $T$, and if $S$-expressions are used in $T$ their semantics is obtained by automatically inserting the implicit morphism.
The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
\end{abstract}

@@

\section{Theories and Theory Morphism}\label{sec:mmt}
\input{mmt}

@@

\section{Implicit Morphisms}
\input{implicit}

\section{Applications}
\input{appl}

\section{Extended Example}
\input{example}

%\section{Complex Theories and Morphisms}\label{sec:complex}
%\input{complex}

\section{Related Work}
\input{related}

\section{Conclusion}
\input{conc}

>>>>

\usepackage{comment}
% flip these to add module expressions; if so, some manual changes to intro and grammars are also necessary
\excludecomment{modexp}
\includecomment{nomodexp}
% include unions as an example of module expressions
\excludecomment{union}

@@

\usepackage{macros/sembracket}
\usepackage{macros/mytikz}
\usepackage{macros/basics}
\usepackage{local}

@@

\begin{abstract}
We introduce \emph{implicit} morphisms as a concept in formal systems based on theories and theory morphisms.
The idea is that there may be at most one implicit morphism from a theory $S$ to a theory $T$, and if $S$-expressions are used in $T$ their semantics is obtained by automatically inserting the implicit morphism.
The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
Concrete applications include elegantly identifying isomorphic theories and extending theories with definitions and theorems as well as efficiently building and maintaining large fine-granular import hierarchies of theories.
\end{abstract}

@@

\section{Theories and Theory Morphisms}\label{sec:mmt}
\input{mmt}

@@

\section{Implicit Morphisms}\label{sec:impl}
\input{implicit}

\section{Applications}\label{sec:appl}
\input{appl}

\begin{modexp}
\section{Advanced Module Expressions}\label{sec:complex}
\input{complex}
\end{modexp}

\section{Conclusion and Related Work}\label{sec:conc}
\input{conc}
